---
title: "An Introduction to Databases"
author: "Zach"
date: "2020-04-13"
output: html_document
---
    
    
```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

## Session Overview
This week as part of the series, Kyle introduced us to some basic database management
by moving the csv data that we dowloaded from the [JHU CSSE database](https://github.com/CSSEGISandData/COVID-19) 
to a local PostgreSQL instance on his machine and demoing some simple commands.  The main 
updates to the code in the session were:

* Instead of downloading every csv from the John's Hopkins database every time we
run the script, only the most recent csv files (i.e. the ones missing from the `data/` directory) are downloaded, and they are automatically inserted into the database hosted on Kyle's work computer ([*insert_report_data.R*](https://github.com/H4estu/COVID19-Data-Exploration/blob/master/scripts/R/session2/insert_report_data.R)).
* A dummy dataset for folks to test out SQL commands ([*simple_database_queries.R*](https://github.com/H4estu/COVID19-Data-Exploration/blob/master/scripts/R/session2/simple_database_queries.R)).
* A basic example of generating a map of the COVID-19 dataset using the `sf` package in R ([*sf_example_simple.R*](https://github.com/H4estu/COVID19-Data-Exploration/blob/master/scripts/R/session2/sf_example_simple.R)).

Check out the [scripts in our GitHub repository](https://github.com/H4estu/COVID19-Data-Exploration/tree/Rmarkdown/scripts/R/session2)
to see everything we did.  

## Database-ics
Put simply, a database is a collection of entities that are stored together.
Commonly, these entities are tables of data, such as the ones you might see
when you open up an Excel spreadsheet.  A very simple example of a one-table database
of carbon dioxide uptake in grass plants data is shown below.

```{r echo=FALSE, fig.fullwidth=TRUE}
library(magrittr)

cap.label <- 'Visualization of a simple database with one table.'
knitr::kable(CO2%>%dplyr::sample_n(.,10)%>%dplyr::mutate(id=seq(1:nrow(.))), caption=cap.label)
```

A key feature of databases are table keys (pun intended).  **Primary keys**, for instance, 
help to organize the data so that it may be efficiently queried.  They also help
to handle bad or null values in your table.  If you have more than one table
in your database, **foreign keys** relate, or connect, the tables in your database
to one another.  

In the `CO2` example above, we would use the `id` column as the primary key.  Since
there are no other tables in the database, we have no need for a foreign key.  You 
can explore this dataset further in R using `datasets::CO2`.


PostgreSQL is one type of relational database management system (RDBMS), and
it is the one we are using for the COVID-19 dataset, as well as for our various projects at work.  

### Database Benefits 
A benefit of using a database management system (DBMS) over other forms of data storage (such as an excel 
or csv file that lives on someone's hard drive), is that DBMSs provide safe 
and consistent means of accessing and manipulating the data.  Many DBMSs (such as PostgresSQL) achieive 
this through what is known as *ACID-compliance*.  I won't go into the details here,
but adherence to these guidelines resolves many of the problems we used to have with
the parcel control sheet, such as people trying to read-from and write-to the control
sheet simultaneously as well as handling an ever-growing list of digitized parcels.  I have
linked an article on ACID-compliant databases at the end of this post if you care
to read further on the topic.

### Some basic commands
Here are some basic SQL commands and their functionalities:
```{r echo=FALSE}
library(data.table)

defs.list <- list('Fetch data from the database (i.e. query the database)',
                  'Edit existing rows in a table',
                  'Add new rows to a table',
                  'Delete rows from a table'
)

defs.dt <- data.table('Command'=list('SELECT','UPDATE','INSERT','DELETE'),
                      'Definition'=defs.list)
knitr::kable(defs.dt)
```

Use these commands to play around with the [dummy data provided by Kyle](https://github.com/H4estu/COVID19-Data-Exploration/blob/master/scripts/R/session2/simple_database_queries.R).


There are other important features of databases such as relations, keys, and schemas
that we won't go into detail here.  However there are plenty of resources online if you
are interested in learning more!


## A Simple Map Using `sf`
The last thing we did in our previous session was to create a global map of the the
COVID-19 dataset.  Each row in the dataset is represented as a point on the map.
Below I've made a map of the cases in the continguious U.S.  I encourage you to 
check out the code in the file [*sf_example_simple.R*](https://github.com/H4estu/COVID19-Data-Exploration/blob/master/scripts/R/session2/sf_example_simple.R) and play around with it yourself!  Note: you will have to be connected to the VPN to access Kyle's computer.

### Connect to the database
```{r connect-database, results='hide'}
library(sf)
library(magrittr)
library(data.table)
library(RPostgreSQL)
library(ggplot2)

git.path <- Sys.getenv('HOME')

source(file.path(git.path,'Code/config_files/db_config.R'))
con <- db_connect.fn()
report_data.dt<-dbGetQuery(con,'SELECT * FROM covid_data.report_data') %>% data.table
dbDisconnect(con)  
# ----------------------------------------------- #
```

### Filter to U.S. cases and map it
```{r sf-example}
# remove NA's 
report_data.dt <- report_data.dt[!is.na(report_data.dt$latitude) & 
                                 !is.na(report_data.dt$longitude),]

# Create spatial object from data table and view on global map
report_data.sf <- st_as_sf(report_data.dt, coords = c("longitude", "latitude"), crs=4326)

us_data.dt <- report_data.dt[report_data.dt$country_region=='US',]
# us_data.dt <- us_data.dt[us_data.dt$confirmed<=sd(us_data.dt$confirmed)]
us_data.dt <- us_data.dt[us_data.dt$confirmed<=1000,]
us_data.sf <- st_as_sf(us_data.dt, coords = c("longitude", "latitude"), crs=4326)

bound <- st_bbox(c(xmin=-125.77148,ymin=23.24135,xmax=-64.59961, ymax=48.86471))

plot(us_data.sf['confirmed'], extent=bound)
```

If the above image does not render for you, you can view the [image here](https://github.com/H4estu/COVID19-Data-Exploration/blob/RMarkdown/blog/content/post/session2/US_48_map.PNG).


There are other important features of databases such as relations, keys, and schemas
that we won't go into detail here.  However there are plenty of resources online if you
are interested in learning more!

***


## Further reading
See the following links for more information on the following topics:

* [**Codecademy article on databases**](https://www.codecademy.com/articles/what-is-rdbms-sql).
Codecademy is also a great place for you to get started with a number of programming languages
on your own with simple introductory lessons for FREE.

* [**ACID-compliant databases**](https://database.guide/what-is-acid-in-databases/) -- ensuring
reliable transactions

* [**sf**](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html) -- package for
creating simple features geometries in R.  This is an excellent way in R to create maps
of geospatial vector data.  We use many aspects of sf in the production code for various projects
to do geospatial operations such as intersections and unions.

